Fractional Types

### 2013-02-04 frac.tex
fix some stuff up, and add some comments here and there

\newcommand{\todo}[1]{\textbf{TODO:} #1}

@@

We are witnessing a convergence of ideas from several distinct research
communities (physics, mathematics, and computer science) towards replacing
\emph{equalities} by \emph{isomorphisms}. The combined programme originating
from these research areas has sparked a huge amount of research that unveiled
new and surprising connections between geometry, algebra, logic, and
computation.

In the physics community, Landauer, Feynman, and others viewed the laws of
physics as fundamentally related to computation. In more detail, the great
majority of the laws of physics are formulated as equalities between
different physical observables. For instance, Newton's second law of
classical mechanics equates the force acting on a system to its rate of
change of momentum. The insight of Landauer, Feynman, etc. is that
\emph{different} physical observables should not be related by an
\emph{equality} but more appropriately by an \emph{isomorphism} that
witnesses, explains, and simulates the process of transforming one observable
to the other.

In the mathematics and logic community, Martin-L\"of developed an extension
of the simply typed $\lambda$-calculus originally intended to provide a
rigorous framework for constructive mathematics. It is now natural to extend
this type theory with a new \emph{identity type} representing the proposition
that two terms are ``equal.'' Specifically, given two terms $a$ and $b$ of
the same type $A$, one forms the type $\texttt{Id}_A(a,b)$ representing the
proposition that $a$ and $b$ are equal: in other words, a term of type
$\texttt{Id}_A(a,b)$ witnesses, explains, and simulates the process of
transforming $a$ to $b$ and vice-versa.

@@

We address this higher-order extension in our paper using a simple novel type
construction, that of a fractional type. The plan of the paper is therefore
as follows: \todo{roadmap}

@@

Note how the above has two readings: first as a set of typing relations for a
set of constants. If these axioms are seen as universally quantified,
orientable statements, they also induce transformations of the (traditional)
values.  The (categorical) intuition here is that these axioms have
computational content because they witness isomorphisms rather than merely
stating an extensional equality.

@@

\begin{definition}{(Syntax of {{langRev}})}
\label{def:Pi}
We collect our types, values, and isomorphisms, to get the full language
definition.
%subcode{bnf} include main
% value types, b ::= 1 | b+b | b*b 
% values, v ::= () | left v | right v | (v,v) 
%
% iso.~types, t ::= b <-> b
% base iso ::= swap+ | assocl+ | assocr+ 
%     &|& unite | uniti | swap* | assocl* | assocr* 
%     &|& dist | factor 
% iso comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
% top level term, l ::= c v
\end{definition}

Mathematically, {{langRev}} can be interpreted in the category of finite sets
and bijections: each type denotes a finite set of the corresponding size and
each combinator {{c : b1 <-> b2}} denotes a bijection between the sets
denoted by {{b1}} and {{b2}}. Operationally, the semantics of {{langRev}} is
given using two mutually recursive interpreters: one going forward and one
going backwards. The use of {{sym}} switches control from one evaluator to
the other. We will not present the operational semantics until the next
section in the context of the extended language. For now, we state without
proof that the evaluation of well-typed combinators always terminates and
that {{langRev}} is logically reversible, i.e., that for all combinators 
{{c : b1 <-> b2}} and values {{v1 : b1}} and {{v2 : b2}} we have the forward
evaluation of {{c v1}} produces {{v2}} iff the backwards evaluation of 
{{c v2}} produces {{v1}}.

@@

@@

\begin{definition}[Syntax of {{langRevTF}}]
\label{def:langRevT}
We extend Def.~\ref{def:Pi} as follows:
%subcode{bnf} include main
% value types, b ::= ... | 1/b
% values, v ::= ... | 1/v
%
% base iso ::= ... | eta*_b | eps*_b
\end{definition}

For a given type {{b}}, the values of type {{1/b}} are of the form {{1/v}}
where {{v : b}}. We think of a value like {{1/left ()}} as a
\emph{first-class constraint} that can only be satisfied if it is matched
with an actual value {{left ()}} or as representing the absence of some
information (i.e., \emph{negative information}), that can only be supplied by
{{left ()}}. The combinator {{eta*_b : 1 <-> 1/b * b}} represents a fission
point which creates --- out of nothing --- an equal amount of negative and
positive information. Symmetrically, the combinator {{eps*_b : 1/b * b <-> 1}}
matches up an equal amount of negative and positive information producing 
no residual information.

>>>>

\newcommand{\todo}[1]{\textbf{TODO:} #1}
\newcommand{\jacques}[1]{\textsc{Jacques says:} #1}

\hyphenation{a-reas}

@@

We are witnessing a convergence of ideas from several distinct research
communities (physics, mathematics, and computer science) towards replacing
\emph{equalities} by \emph{isomorphisms}. The combined programme originating
from these research areas has sparked a huge amount of research that unveiled
new and surprising connections between geometry, algebra, logic, and
computation (see \cite{baez2011physics} for an overview).

In the physics community, Landauer~\cite{Landauer:1961}, Feynman, and 
others have interpreted the laws of
physics as fundamentally related to computation. In more detail, the great
majority of the laws of physics are formulated as equalities between
different physical observables. For instance, Newton's second law of
classical mechanics equates the force acting on a system to its rate of
change of momentum. The insight of Landauer, Feynman, etc. is that
\emph{different} physical observables should not be related by an
\emph{equality} but more appropriately by an \emph{isomorphism} that
witnesses, explains, and models the process of transforming one observable
to the other.

In the mathematics and logic community, Martin-L\"of developed an extension
of the simply typed $\lambda$-calculus originally intended to provide a
rigorous framework for constructive mathematics. This has been further
extended (first by Streicher and Awodey, and subsequently by many others)
with a new \emph{identity type} representing the proposition
that two terms are ``equal.'' Specifically, given two terms $a$ and $b$ of
the same type $A$, one forms the type $\texttt{Id}_A(a,b)$ representing the
proposition that $a$ and $b$ are equal: in other words, a term of type
$\texttt{Id}_A(a,b)$ witnesses, explains, and models the process of
transforming $a$ to $b$ and vice-versa.

@@

This paper addresses this higher-order extension.  It does this via a novel
type construction, which we dub a \emph{fractional type} because of its
interaction with product types.  By adding a formal (linear) dual to
(type theoretic) products, we can internalize the notion that ``taking an
input'' corresponds to negative information, which turns out to be the 
basis of first-class functions.

The plan of the paper is therefore
as follows: \todo{roadmap}

@@

Note how the above has two readings: first as a set of typing relations for a
set of constants. Second, if these axioms are seen as universally quantified,
orientable statements, they also induce transformations of the (traditional)
values.  The (categorical) intuition here is that these axioms have
computational content because they witness isomorphisms rather than merely
stating an extensional equality.

@@

\begin{definition}{(Syntax of {{langRev}})}
\label{def:Pi}
%subcode{bnf} include main
% value types, b ::= 1 | b+b | b*b 
% values, v ::= () | left v | right v | (v,v) 
%
% iso.~types, t ::= b <-> b
% base iso ::= swap+ | assocl+ | assocr+ 
%     &|& unite | uniti | swap* | assocl* | assocr* 
%     &|& dist | factor 
% iso comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
% top level term, l ::= c v
\end{definition}

Mathematically, {{langRev}} can be interpreted in the category of finite sets
and bijections: each value type denotes a finite set of the corresponding size
and each combinator {{c : b1 <-> b2}} denotes a bijection between the sets
denoted by {{b1}} and {{b2}}. Operationally, the semantics of {{langRev}} is
given using two mutually recursive interpreters: one going forward and one
going backwards. The use of {{sym}} switches control from one evaluator to
the other. We will not present the operational semantics until the next
section, in the context of the extended language. For now, we state without
proof that the evaluation of well-typed combinators always terminates and
that {{langRev}} is logically reversible, i.e., that for all combinators 
{{c : b1 <-> b2}} and values {{v1 : b1}} and {{v2 : b2}} we have the forward
evaluation of {{c v1}} produces {{v2}} iff the backwards evaluation of 
{{c v2}} produces {{v1}}.

\jacques{Since we have Agda code for the above, should we refer to it
here, and make it available?}

Note that a small modification of {{lanRev}} (the addition of an
empty type {{0}} with its usual laws) would exactly capture, at the type
level, the notion of \emph{semiring} (occasionally called a \emph{rig}) where
we replace equality by isomorphism.  This can also be also compared to
\emph{bimonoidal categories}.  As is usual with with type theories, the
coherence conditions (pentagonal and triangle identity) are not explicit,
but would correspond to certain identity types being trivial.  {{lanRev}}
instead, at the type leve, models a commutative ringoid where the
multiplicative structure forms a commutative monoid, but the additive structure
is just a commutative magma.  We chose not to include {{0}} as a technical
convenience for the next section.

@@

{{langRev}} models isomorphisms of values rather well.  However, purposefully,
it has no notion of \emph{input} or \emph{output}; more precisely, because
of strict preservation of information, these two concepts coincide.  This
is the root of reversibility.

But if we want to model functions of any flavor, we need to differentiate
between these notions.  Our idea, inspired by computational dualities and
polarity, is to introduce a \emph{formal dual} for our types.  We thus
consider a type in a negative position in a type to possess negative
information, as it is really a request for information.  Since information
is logarithmic, this means that a request for information should behave
like a \emph{fractional type}.

@@

\begin{definition}[Syntax of {{langRevTF}}]
\label{def:langRevT}
We extend Def.~\ref{def:Pi} with:
%subcode{bnf} include main
% value types, b ::= ... | 1/b
% values, v ::= ... | 1/v
%
% base iso ::= ... | eta*_b | eps*_b
\end{definition}

For a given type {{b}}, the values of type {{1/b}} are of the form {{1/v}}
where {{v : b}}. Note that {{1/v}} is purely formal, as is {{1/b}}.
However, we think of a value like {{1/left ()}} as a
\emph{first-class constraint} that can only be satisfied if it is matched
with an actual value {{left ()}} or as representing the absence of some
information (i.e., \emph{negative information}), that can only be supplied by
{{left ()}}. The combinator {{eta*_b : 1 <-> 1/b * b}} represents a fission
point which creates --- out of no information --- an equal amount of negative
and positive information. Symmetrically, the combinator {{eps*_b : 1/b * b <->
1}} matches up an equal amount of negative and positive information producing 
no residual information.  The best analogy is probably that of vectors and
covectors, also know as column vectors and row vectors (respectively): in
the finite dimensional case, they are isomorphic, but they are nevertheless
quite different, being $(1,0)$ and $(0,1)$ tensors (respectively).  In
other words, since we are explicitly interested in \emph{isomorphisms}, 
we should not conflate negative-information values with positive-information
values just because they happen to be isomorphic.
